---
title: "Getting started with manifest"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with manifest}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## Why keep a manifest?

Before a dataset reaches a model it is usually *sampled*: outliers are trimmed,
a geography or time window is chosen, invalid records are removed. Each of these
decisions changes who is in the analysis and who is not, yet the reasoning
behind them tends to live only in the analyst's head or in scattered code
comments. Months later — during peer review, replication, or a hand-off — it can
be surprisingly hard to answer a simple question: *how did we get from the raw
data to the modeling sample, and why?*

The **manifest** package produces a lightweight audit trail for exactly this
stage of the workflow. You bracket your filtering code with `begin_manifest()`
and `end_manifest()`, and the package records, for every row-reducing step:

- a human-readable **rationale** (drawn from `#:` comments in the code),
- the exact **code** that ran,
- the number and percentage of rows **dropped**, and
- the rows **remaining**.

The result is printed as a table and saved as a timestamped CSV that can travel
with your project as documentation of the sampling framework.

## Comment conventions

manifest reads your script and treats comments differently depending on their
prefix:

| Prefix | Meaning              | Captured in the manifest? |
|:-------|:---------------------|:--------------------------|
| `#:`   | Manifest criteria    | **Yes** — becomes the `CRITERIA` text |
| `#: {name} ...` | Step label  | **Yes** — becomes the `STEP` label |
| `##`   | Dev / internal note  | No |
| `#`    | Ordinary comment     | No |

A `#:` comment sitting directly above a filtering call is paired with that call.
This lets you write the *reason* for a decision as a normal comment and have it
promoted automatically into the documentation.

## A worked example

The package ships a complete, runnable example. Here is the script in full — note
the `#: {name}` labels and `#:` rationales above each filtering step:

```{r, echo = FALSE, results = "asis"}
script <- system.file("examples", "example_cleaning.R", package = "manifest")
cat("```r", readLines(script), "```", sep = "\n")
```

Because the example is executed with `source()`, `begin_manifest()`
auto-detects the script's path, so `end_manifest()` can read it back and
back-fill the `CRITERIA` and `CODE` columns from the `#:` comments. Run it:

```{r, message = TRUE}
library(manifest)

script <- system.file("examples", "example_cleaning.R", package = "manifest")
source(script)
```

`end_manifest()` also wrote a timestamped CSV. Reading it back shows the
structured record that would live alongside your project:

```{r}
out_dir  <- file.path(tempdir(), "manifest")
csv_file <- tail(sort(list.files(out_dir, pattern = "\\.csv$", full.names = TRUE)), 1)

knitr::kable(read.csv(csv_file, check.names = FALSE))
```

## Anatomy of the workflow

**`begin_manifest(data, ...)`** opens the log and records the starting row
count. Two arguments are worth calling out:

- `description` — a sentence about the dataset, stored on the `START` row.
- `build_script` — the path to the script that *assembled* the raw data (often
  distinct from the cleaning script you are annotating). It is recorded so the
  manifest points back to the data's origin.

**`mfilter(data, ...)`** is a drop-in replacement for `dplyr::filter()`. It
applies the filter *and* records the row delta. Pair it with a `#:` comment on
the line above to supply the rationale automatically.

**`end_manifest(data)`** parses the script for `#:` comments, back-fills the
`CRITERIA` / `CODE` columns, prints the summary table, and writes the CSV.

**`manifest_log()`** lets you peek at the running log at any point *before*
`end_manifest()`.

### Reading the columns

| Column        | Meaning |
|:--------------|:--------|
| `STEP`        | Step number, plus the `{name}` label if one was given |
| `DROPPED`     | Rows removed by this step |
| `DROPPED_PCT` | Those rows as a share of the **starting** N |
| `REMAIN`      | Rows still in play after the step |
| `REMAIN_PCT`  | Remaining rows as a share of the starting N |
| `CRITERIA`    | The rationale, from the `#:` comment |
| `CODE`        | The exact expression that ran |

Note that percentages are always relative to the original row count, so
`REMAIN_PCT` reads as "share of the raw data still in the modeling sample" — a
convenient headline number for a methods section.

## Steps that aren't simple filters

Not every row-reducing operation is a `filter()`. Joins, de-duplication, and
aggregation also drop rows, and for those you can log the step manually with
`log_step()`, passing the rationale explicitly via `criteria =`:

```{r}
library(dplyr)

set.seed(1)
roster <- data.frame(
  id     = 1:20,
  region = sample(c("A", "B"), 20, replace = TRUE),
  active = sample(c(TRUE, FALSE), 20, replace = TRUE)
)

begin_manifest(
  roster,
  description  = "Toy roster to demonstrate log_step()",
  output_dir   = file.path(tempdir(), "roster-manifest"),
  dataset_name = "roster",
  verbose      = FALSE
)

# A normal filter, tracked automatically:
roster <- mfilter(roster, active)

# A de-duplication that filter() can't express — log it by hand:
roster <- roster[!duplicated(roster$region), ]
roster <- log_step(roster, criteria = "Keep one representative record per region")

manifest_log()
```

Here `mfilter()` and `log_step()` interleave freely, and `log_step()` carries
its own `criteria` string rather than relying on a `#:` comment. Call
`end_manifest()` when you are done to finalize the table and write the CSV, just
as in the first example.

## Where to go next

- Run the shipped example yourself:
  `source(system.file("examples", "example_cleaning.R", package = "manifest"))`
- See `?begin_manifest`, `?mfilter`, `?log_step`, and `?end_manifest` for the
  full argument reference.
